---
title: "DE Analysis Report (EV Proteomics): `r params$title`"
subtitle: "`r params$subtitle`"
author: "`r params$author`"
format: 
  html:
    fig-width: 12
    fig-height: 10
    toc: true
    toc-title: 'DE Analysis'
    title-block-banner: '#1b2944'
    title-block-banner-color: '#f79646'
    lightbox: auto
execute: 
  echo: false
  warning: false
embed-resources: true
theme: 
 - layout/brand 
 - layout/custom.scss

params:
  description: ""
  title: ""
  subtitle: ""
  author: "" 
  formula: ""
  comparisons: ""
  contrast: ""
  input_file: ""
  design_file: ""
  FC_thr: 1
  adjpval_thr: 0.05
  Proteotypic: TRUE
  pep_per_prot: 3
  nNonZero: 30
  aggr_method: "medianPolish"
  normalization: "quantiles"
  filtering_contaminant: ""
  contaminant_str: ""
  folder_prj: ""
  ensembl_annotation: ""
  ensembl_col: ""
  confounder_list: ""
  PCA_comparison: "Group"
  quantitative_features: "Precursor.Quantity"
  filtPerGroup: ''
  mbr: FALSE
  wildstr_run: "CMB-"
  DIANN_ver2: FALSE 
  keep_design_order: FALSE
  comparison_label: ""
  qf_obj: ""
  de_obj: ""
  filt_NaNDE: ""
  part_item: ""
  part_value: ""
---

```{r input}
 #| output: false
#source('utils_function.R')

 required_packages = c('QFeatures',"ggrepel","tidyr","tibble",
                       "plotly","DT","MSnbase",'gridExtra','dplyr','stringr' ,"UpSetR",'logger','pander','heatmaply','GGally','yaml','webshot2','htmlwidgets')
# #print(required_packages)
 diareport::check_dependencies(required_packages)

```

## Experiment Description

`r params$description`

```{r check_parameter}
# 
# log_threshold(INFO)
log_appender(appender_console)
log_appender(appender_file(file.path( params$folder_prj, "logfile_protein.log")), index = 2)
# formatter_data_frame <- function(df, ...) {
#      pander::pander_return(df, style = 'simple')
# }
# log_formatter(formatter_data_frame)
 log_formatter(formatter_glue)


check_and_substitute_forbidden_chars <- function(input_string) {
  # Define forbidden characters for filenames in Windows
  forbidden_chars <- c("<", ">", ":", "\"", "/", "\\", "|", "?", "*")

  # Iterate over each forbidden character and replace it with an empty space
  for (char in forbidden_chars) {
    input_string <- gsub(char, " ", input_string, fixed = TRUE)
  }
  return(input_string)
}
## present two times , but not exported 
is_empty <- function(x) {
    is.character(x) && length(x) == 1 && x == ''
  }

theme_custom_vis <- function(base_size = 12) {
  theme_bw(base_size = base_size) %+replace%
    theme(
      # leggend
      legend.title = element_text(size = rel(0.85), face = "bold"),
      legend.text = element_text(size = rel(0.70), face = "bold"),
      legend.key = element_rect(fill = "transparent", colour = NA),
      legend.key.size = unit(1.5, "lines"),
      legend.background = element_rect(fill = "transparent", colour = NA),
      strip.background = element_rect(fill = "#7c7c7c", color = "#7c7c7c"),
      strip.text = element_text(size = rel(0.85), face = "bold", color = "#1b2944", margin = margin(5,0,5,0))
    )
}


 
if (params$filtPerGroup == ''){
      label_filtering <- ' among all the samples.'
      }
if (params$filtPerGroup == 'all'){
      label_filtering <- ' in each group.'
      }
if (params$filtPerGroup == 'at_least_one'){
      label_filtering <- ' in at least one group.'
}


# make a subfolder for each contrast
if (!dir.exists(file.path(params$folder_prj))){
  dir.create(file.path( params$folder_prj),recursive = TRUE)
}
 dir.create(file.path( params$folder_prj, "Result"),recursive = TRUE)

 theme_set(theme_custom_vis(base_size = 18))
 
 st=format(Sys.time(), "%Y%m%d_%H%M")

write(as.yaml(params), file =  file.path( params$folder_prj, paste0("DIAReport_parameter_",st  , ".yaml")))

```

## Import data

Reading DIA data from **`r basename(params$input_file)`** using *`r params$quantitative_features`* from DIA-NN as quantitative features.

`r if (! params$mbr) 'All the precursors with *Global Q-value and Global PG Q-value < 0.01* are used for the analysis' else 'All the precursors with *Lib Q-value and Lib PG Qvalue < 0.01* are used for the analysis'`

```{r import data}


pe <- readRDS(params$qf_obj )

res_de <- readRDS(params$de_obj )
res_part_item  <-  readRDS(params$part_item )
res_part_value  <- readRDS(params$part_value )

L <- readLines(params$design_file, n = 1)
if (grepl(";", L)) design <- read.csv2(params$design_file) else design <- read.csv( params$design_file)

```

## Filtering steps

The filtering steps applied are:

-   Keeping only precursors with more than \>= **`r params$nNonZero`%** of *non missing* values, `r label_filtering`
-   `r if (params$Proteotypic) 'Selecting only proteotypic peptides' else 'Selecting both not and proteotypic peptides'`
-   Include proteins with at least **`r params$pep_per_prot` peptides**
-   `r if (params$filtering_contaminant) 'Contaminant proteins are removed ' else 'Contaminant proteins are not filtered'`

```{r Confanalysis_title, results='asis'}

if ( ! is_empty(params$confounder_list)  )  {
  
  header <- "Confounder Analysis"
  
  cat('##', header, '\n\n')
  
  
}

```

```{r confounder_value_analysis}
#| layout-ncol: 2
#| layout-nrow: 2
#| lightbox: true
log_info('Confounder Analysis')
cData<-as.data.frame(colData(pe))
list_numeric <- list()
list_character <- list()

if ( ! is_empty(params$confounder_list) ) {
  for  (col in params$confounder_list){
     if  (is.factor (cData[[col]])) {
       #log_info(paste0('Character - ', col))
       list_character <- append(list_character, col)
        
     }else {
       #log_info(paste0('Numeric - ', col))
       list_numeric <- append(list_numeric, col)
     }  
  }

  if (length(list_numeric) != 0) {
  
    cData_n<-cData%>%pivot_longer(cols=unlist(list_numeric), names_to="Numeric_Variables", values_to="Value")

		           boxplot_num <-ggplot(cData_n, aes(x=Numeric_Variables, y=Value, fill=Group))+geom_boxplot()
  print(boxplot_num)
               pdf(file= file.path(params$folder_prj,"Result",paste0('Boxplot_num_confounder.pdf')),paper="a4")
               boxplot_num <-ggplot(cData_n, aes(x=Numeric_Variables, y=Value, fill=Group))+geom_boxplot()
  print(boxplot_num)
      invisible(dev.off())
  } 
  
  for (col_ch in unlist(list_character)){
    #log_info(col_ch)
    boxplot_ch<- ggplot(cData, aes(x={{col_ch}}, fill=Group))+geom_bar()+ggtitle( {{col_ch}} )
    print(boxplot_ch)
          pdf(file= file.path(params$folder_prj,"Result",paste0('Boxplot_ch_confounder.pdf')),paper="a4")
         boxplot_ch<- ggplot(cData, aes(x={{col_ch}}, fill=Group))+geom_bar()+ggtitle( {{col_ch}} )
    print(boxplot_ch)  
     
      invisible(dev.off())
  }

}


```

## Contaminant Analysis

```{r contaminat_visual}
#| layout-ncol: 2
#| layout-nrow: 1
#| lightbox: true

# df <- rowData(pe[['proteinRS_']]) %>% as.data.frame() %>% select( Protein.Ids, Protein.Names)
# df <- df %>% mutate( type =sapply(
#         str_split(Protein.Names, "_"), 
#         function(x) {
#           # Take the second part, split by semicolon if present
#           second_part <- x[2]
#           if (is.na(second_part)) return(NA)
#           # If there's a semicolon, split and check for BOVIN
#           splitted <- unlist(str_split(second_part, ";"))
#           if ("BOVIN" %in% splitted) {
#             return("BOVIN")
#           } else { 
#             return("HUMAN")
#           }
#         }
#     ))
# count_prot <- df %>% dplyr::count(type)


df <- rowData(pe[['precursor_']]) %>% as.data.frame() %>% select( Protein.Ids, Protein.Names)
df <- df %>% mutate( type =sapply(
        str_split(Protein.Names, "_"), 
        function(x) {
          # Take the second part, split by semicolon if present
          second_part <- x[2]
          if (is.na(second_part)) return(NA)
          # If there's a semicolon, split and check for BOVIN
          splitted <- unlist(str_split(second_part, ";"))
          if ( ! (  'HUMAN' %in% splitted)) {
            return("CONTAMINANT")
          } else { 
            return("HUMAN")
          }
        }
    )
  )
count_precursor <- df %>% dplyr::count(type)


## get the list of the precursor
search_map <- c("HUMAN" = "HUMAN", "CONTAMINANT" = params$contaminant_str)
type_names <- names(search_map) # This gives you c("HUMAN", "CONTAMINANT")


 #type_names <- c('CONTAMINANT','HUMAN')
  
  # Create a named list of Precursor.Id vectors, one per tissue
  precursor_lists <- lapply(type_names, function(type) {
    as.data.frame(rowData(pe[['precursor_']])) %>%
      filter(str_detect (Protein.Names ,search_map[type] )) %>%
      pull(Precursor.Id)
  })
  
    names(precursor_lists) <- type_names  

## relative % 
  total_ammount <- assay(pe[['precursor_']]) %>% base::colSums(na.rm = TRUE)
  
  d <- data.frame(
  Sample = names(assay(pe[['precursor_']])[precursor_lists[['CONTAMINANT']], ] %>% base::colSums(na.rm=TRUE))
)
 
  for (a in names(precursor_lists)){
    
      d[a] <- (assay(pe[['precursor_']])[precursor_lists[[a]],] %>% base::colSums(na.rm=TRUE) / total_ammount ) * 100 

  }
 log_info(paste( dim(d), collapse = ' '))
 d_visual <- d %>% tidyr::gather( Contaminant , percentage ,- Sample)

# cont_proteinID <- ggplot(count_prot, aes(x = "", y = n, fill = type)) +
#   geom_bar(stat = "identity") +
#   labs(y = "# Protein", x = "", fill = "Type") 
# 
# print(cont_proteinID)

cont_precursor <- ggplot(count_precursor, aes(x = "", y = n, fill = type)) +
  geom_bar(stat = "identity") +
   scale_fill_manual(values = c(
    "CONTAMINANT" = "#F8766D",
    "HUMAN" = "#00BFC4"
  ))+
  labs(y = "# Precursor", x = "", fill = "Type") 

print(cont_precursor)

plot_perc <- ggplot( d_visual, aes(x= Sample, y = percentage ,fill = Contaminant ) ) +  
  geom_col(position = "stack") +
   scale_fill_manual(values = c(
    "CONTAMINANT" = "#F8766D",
    "HUMAN" = "#00BFC4"))+ 
   scale_x_discrete(guide = guide_axis(angle = 90) )  + ylab("percentage of precursors") 

ggsave(file.path(params$folder_prj,"Result","EV_contaminant_percentage_precursor_intensity.pdf"), 
       plot = plot_perc, width = 8.27, height = 11.69, units = "in")

print(plot_perc)
```

## Normalization

The precursor intensities are:

-   Log2 transformed and then normalized using the **`r params$normalization`** approach.
-   Precursors/peptides are summarized at protein level using the **`r params$aggr_method`** approach.

After these steps, the quantitative features used in the downstream analysis are:

-   Number of precursors identified/quantified is : **`{r} dim(assay(pe[['precursor']]))[1]`**
-   Number of proteins identified/quantified is : **`{r} dim(assay(pe[['proteinRS']]))[1]`**

```{r normalization_plot}
#| lightbox: false
if ( 'precursorLog' %in% names(pe)){
  par(mfrow=c(1,3))
limma::plotDensities(assay(pe[["precursorLog"]]),legend=FALSE,main='Log2-Transform')
limma::plotDensities(assay(pe[["precursorNorm"]]),legend=FALSE,main='Normalized') 
limma::plotDensities(assay(pe[["proteinRS"]]),legend=FALSE,main='Protein Summarization' )
}else{
   par(mfrow=c(1,2))
  limma::plotDensities(assay(pe[["precursorNorm"]]),legend=FALSE,main='Normalized') 
limma::plotDensities(assay(pe[["proteinRS"]]),legend=FALSE,main='Protein Summarization' )
}


```

## Data Missing Analysis

The plot shows the completeness of the experiments at precursor and summarized level. On the the x-axis peptides/proteins are ranked based on the increasing NA content.

```{r missing_val_analysis}
#| layout-ncol: 2
#| layout-nrow: 1
#| lightbox: true


pdf(NULL) 
  peptidemissingness <- MSnbase::plotNA(assay(pe[["precursorNorm"]]))
invisible(dev.off())

peptidemissingness <-  peptidemissingness + 
                         xlab("Precursor index (ranked by data completeness)") +
                           ggtitle("Missingness - Precursor level" )

peptidemissingness


pdf(NULL) 
  proteinmissingness <- MSnbase::plotNA(assay(pe[["proteinRS"]]))
invisible(dev.off())
#proteinmissingness <- MSnbase::plotNA(assay(pe[["proteinRS"]])) 

proteinmissingness <- proteinmissingness +
     xlab("Protein index (ranked by data completeness)") +
     ggtitle("Missingness - Protein Level")

proteinmissingness

```

In the following table, total number of and percentace of missing protein values are showed for all samples.

```{r %_miss_sample}

DT::datatable( as.data.frame(nNA(pe[["proteinRS"]])$nNAcols) %>% rename(  Sample_Name = name) %>% arrange(desc(pNA))  ,
              extensions = c('FixedColumns', 'Scroller'),
              options = list(fixedColumns = TRUE, scrollY = 400, scrollX = TRUE,
                             scroller = TRUE, dom = 'Bfrtip',
                             autoWidth = TRUE
                             ),
              ) %>% 
  formatPercentage(columns = 'pNA', digits = 2) # Multiplies by 100 and adds %

```

## Completeness Analysis

This section analyzes the completeness of protein readouts across all samples within each group. For each quantified protein, a completeness percentage is calculated based on the proportion of missing values across all samples in the group. These percentages are compared between the groups in the lower diagonal. On the diagonal, a histogram illustrates the distribution of detection percentages for each group.

```{r completeness_analysis}
#| lightbox: false

log_info('Completness Analysis...')
   custom_lower <- function(data, mapping, ...) {
    
      x_data <- data[[as_label(mapping$x)]]
      y_data <- data[[as_label(mapping$x)]]
     
      
      # Calculate mean and standard deviation
      mean_x <- mean(x_data, na.rm = TRUE)
      mean_y <- mean(y_data, na.rm = TRUE)
      sd_x <- sd(x_data, na.rm = TRUE)
      sd_y <- sd(y_data, na.rm = TRUE)
      
      ggplot(data = data, mapping = mapping) +
      geom_jitter(  size=1,alpha= 0.8)+
      geom_abline(slope = 1, intercept = mean_y - mean_x, color = 'blue')+
      geom_abline(slope = 1, intercept = mean_y - mean_x + sd_y + sd_x, color = 'red', linetype = "dashed") +       # Mean + sigma line
        geom_abline(slope = 1, intercept = mean_y - mean_x - sd_y - sd_x, color = 'red', linetype = "dashed")       # Mean - sigma line
      
      
    }
    
    
    custom_diag <- function(data, mapping, ...) {
      ggplot(data = data, mapping = mapping) + 
        geom_histogram(binwidth = 10, fill = "blue", color = "black") 
    }
    
    
    
     if  (dim(colData(pe) %>% as.data.frame() %>% distinct(Group))[1] <= 6) {  
    
    as.data.frame(rowData(pe[['proteinRS']])) %>% 
      select(matches("^perc"))  %>% 
      ggpairs(
               upper = list(continuous = wrap("blank")),
              diag = list(continuous = custom_diag),
      lower = list(continuous = custom_lower),
      xlab = "percentage of detection",
      ylab = "percentage of detection"
      )}else{       
         cat('Too many groups detected (> 6) for the completness analysis plot')
      }
  

```

## PCA

```{r pca_plot}
#| results: asis

# to do make in for loop for a set of variable
log_info('PCA Analysis...')

res_pca  <- diareport::generate_pca_plots( pe, params, layer ='proteinRS' )

if (any (res_pca$status == 1) ){
  log_info(paste('PCA plots :  one plot will be skipped: ', res_pca$msgs[ which(res_pca$status==1)]))
}
l_pca <- res_pca$plots 

plotly_list <- lapply(l_pca, function(p) if (!is.null(p)) ggplotly(p))
# Remove NULLs (from failed/skipped PCAs)
plotly_list <- Filter(Negate(is.null), plotly_list)


htmltools::tagList(plotly_list)

```

## EV Protein Marker

In this section, the protein intensities of 7 EV marker proteins (CD9, CD63, SDCBP, CD8, TSG101, PDCD6IP, FLOT1) are summarized for all groups. The dark dot represents the mean of the sample intensities for each protein, and the length of the bar corresponds to the interval $\mu \pm \sigma$. If some markers are not shown, it means that they are not identified/ quantified in the experiment. 

```{r protein_marker}
#| results: asis

marker <- read.csv('ev_marker.tsv', sep=';')	
p_m <- NULL 

inte_marker <- intersect(marker$Accession , rownames(assay(pe[['proteinRS']])))
miss_marker <- setdiff( marker$Accession , rownames(assay(pe[['proteinRS']]))   ) 
#log_info(length(miss_marker))
#log_info( paste(inte_marker, collapse  =' '))
if (length(miss_marker) == 8){
  cat( 'None of EV markers presented in the database have been detected')
}else{
  log_info( paste(inte_marker, collapse  =' '))

   ## just the the intersection. 
   f <- assay(pe[['proteinRS']])[inte_marker , ] %>% as.data.frame() %>% 
  tibble::rownames_to_column('protein') %>% 
  left_join( marker %>% select(Accession, Gene) , join_by(protein == Accession )) %>% 
  tidyr::gather( Sample , summarize_intensities ,- Gene, -protein) %>%   
  left_join( colData(pe) %>%  as.data.frame() %>% select(SampleName , Group), join_by(Sample == SampleName ))
  log_info( paste(dim(f), collapse = ''))

    p_m <- ggplot(f, aes(x = Group, y = summarize_intensities, color = Gene, group = Gene)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.15), alpha = 0.3, size = 2) +
  stat_summary(
    fun = mean,
    fun.min = function(x) mean(x) - sd(x),
    fun.max = function(x) mean(x) + sd(x),
    geom = "pointrange",
    position = position_dodge(width = 0.1),
     size = 1) 
    ggsave(file.path(params$folder_prj,"Result","EV_marker.pdf"), 
       plot = p_m, width = 10.27, height = 11.69, units = "in")
  }
l_marker <- list(p_m)
plotly_mark <- lapply(l_marker, function(p) if (!is.null(p)) ggplotly(p))
# Remove NULLs (from failed/skipped PCAs)
plotly_mark <- Filter(Negate(is.null), plotly_mark)
htmltools::tagList(plotly_mark)

```

## Type of Protein

Using an internal annotation database (*ProteinAnnotationsShortList_0.9*), proteins are annotated as intracellular, secreted, transmembrane or other. The number of identified proteins in each category is color-coded by the experimental group.

```{r count_annotation }
#| lightbox: true

ann <- read.csv('ProteinAnnotationsShotList.txt', sep='\t')

 type_names <- c('TM.protein','Intracellular','Secreted','Other')
  
  # Create a named list of Precursor.Id vectors, one per tissue
  proteinID_lists <- lapply(type_names, function(type) {
    ex <- ann %>% filter(Subcellular.Location == type ) %>% pull (Entry)
    ex [ex  %in% rownames(assay(pe[['proteinRS']]))]

  })
  
  names(proteinID_lists) <- type_names  
  #total_ammount <- assay(pe[['precursor_']]) %>% base::colSums(na.rm = TRUE)
  
  d <- data.frame(
  Sample = colData(pe)$SampleName)
 
  for (a in names(proteinID_lists)){
       d[a] <- colSums(!is.na(assay(pe[['proteinRS']])[proteinID_lists[[a]], , drop = FALSE] ))

  }
  
 d <- d %>% left_join(colData(pe)%>% as.data.frame() %>% select(SampleName, Group), join_by( Sample == SampleName) )
 d_count <- d %>% tidyr::gather( Type , N_protein ,- Sample, -Group)
    
 d_sum <- d_count %>%
  group_by(Group, Type) %>%
  summarize(total_proteins = sum(N_protein), .groups = "drop")
    
# p_ann <-ggplot(d_sum, aes(x = Group, y = total_proteins, fill = Type)) +
#   geom_bar(stat = "identity") +
#   labs(x = "Group", y = "Number of Proteins", fill = "Protein Type")
# print(p_ann) 

p_ann_ <-ggplot(d_sum, aes(x = Type, y = total_proteins, fill = Group)) +
  geom_bar(stat = "identity") +
  labs(x = "Type", y = "Number of Proteins", fill = "Group")

print(p_ann_) 

ggsave(file.path(params$folder_prj,"Result","EV_protein_annotated_counts.pdf"), 
       plot = p_ann_, width = 8.27, height = 11.69, units = "in")

```

## DE Analysis

MSqRob2 was used with the following formula `r params$formula` with ridge regression disabled

```{r prepare_save_result_de}

if (!dir.exists(file.path(params$folder_prj, "Result","Contrast"))){
  dir.create(file.path( params$folder_prj, "Result","Contrast"), recursive = TRUE)
}
# old one params$comparisons
for (a in   params$comparison_label ){
  log_info(a)
  log_info(check_and_substitute_forbidden_chars(a))
  dir.create(file.path(params$folder_prj,"Result","Contrast",check_and_substitute_forbidden_chars(a)), recursive = TRUE,showWarnings = TRUE)
}  
log_info('Saving pe object (qfeatures) as Rds file.')
saveRDS(pe, file = file.path( params$folder_prj, "Result","qfeat_experiment.Rds"))  

```

## QC plots

::: panel-tabset
```{r QC_plot}
#| results: asis
  for (cmp in names(res_de)) {
    cmp_label <- gsub(params$contrast,'',cmp )
    cat( '##', paste(cmp_label,' P-values distribution 
                     ',sep=' '), '\n\n')
    
    diareport::render_child(data = res_de[[cmp]], 
                 pe =pe , 
                 label= cmp ,
                 path=file.path(params$folder_prj, "Result","Contrast",check_and_substitute_forbidden_chars(cmp)), template='_templatePval_distribution.Rmd') 
    cat('\n\n')
    
  }

```
:::

## Comparison

::: panel-tabset
```{r comparison}
#| results: asis

log_info('Running Analysis for each comparison..')

  for (cmp in names(res_de)) {
        log_info( paste0('  --  ',cmp)) 

    cmp_label <- gsub(params$contrast,'',cmp )
    cat( '## ', paste(cmp_label,sep=' '), '\n\n')
    cat ('This volcano plot summarizes the differential expression landscape in the comparison between two groups \n')
    
    
    diareport::render_child(data = res_de[[cmp]],
                 pe =pe,
                 label = cmp , 
                 path=file.path(params$folder_prj, "Result","Contrast",check_and_substitute_forbidden_chars(cmp)),  
                 template='_templateContrast_EV.Rmd') 
    cat('\n\n')
    
   

    if ( dim(res_de[[cmp]]$toptable %>% filter(differential_expressed %in% c('UP','DOWN')) )[1] <= 5 || dim(res_de[[cmp]]$toptable %>% filter(differential_expressed %in% c('UP','DOWN')) )[1]  >=  50     ){
      cat(' Not enough DE proteins (< 5) or too many DE proteins for heatmap visualization \n')
    }else{
      
    contrast_values_parsed <- diareport::parse_comparison(cmp, params$contrast,pe)

    filt_val <- diareport::select_samples_comparison(contrast_values_parsed, pe , params$contrast)
    
    diareport::render_child(data = res_de[[cmp]],
                   pe = pe,
                   sample_rel = filt_val,
                 label = cmp,
                 path= file.path(  params$folder_prj, "Result","Contrast",check_and_substitute_forbidden_chars(cmp)),
                 template='_templateHeatMap.Rmd')
    }
    cat('\n\n')
    cat('\n\n')
    cat('###','Absent-from-DE Analysis')
    cat('\n\n')
    cat ('All proteins that are either uniquely present or absent in one of the two groups are listed below. Because these proteins are identified exclusively in one group, they are not included in the standard differential expression (DE) results and are therefore listed separately here. The heatmap displayed above the table illustrates the protein intensities across all replicates for the two groups of interest.\n')
    ### partial analysis result visualition   
    if (dim( res_part_item[[cmp]])[1]==0){
       cat('\n\n')
  cat('No proteins present.')
   cat('\n\n')
      }else{
    # reas_part_item  <-  readRDS(params$part_item )
    #res_part_value  <- readRDS(params$part_value )
    diareport::render_child(data = res_part_item[[cmp]],
                  pe = pe,
                  label = cmp , 
                  sample_rel = res_part_value[[cmp]] ,
                  path=file.path(params$folder_prj, "Result","Contrast",check_and_substitute_forbidden_chars(cmp)),  
                  template='_templatePartialResult.Rmd') 
     cat('\n\n')
    cat('\n\n')
    }
    cat('\n\n')
    cat('\n\n')
    cat('These bar plots summarize the number of significantly upregulated/downregulated  proteins based on different adjusted p-values (selected adjusted p-values are 0.001, 0.01, 0.05, and 0.1 - see facet headers) and log2 fold-change thresholds (on the x-axis) used to define the significance levels.')
   
    diareport::render_child(data = res_de[[cmp]],
                 pe =pe,
                 label = cmp,
                 path= file.path( params$folder_prj, "Result","Contrast",check_and_substitute_forbidden_chars(cmp)),
                 template='_templateBarPlot.Rmd')
    cat('\n\n')
    
    
  }

```
:::

## Summary of the DE proteins

The table shows a summary of the number of DE proteins found in all comparisons. The overlap of DE proteins among the comparisons is also showed using an upset plot.

```{r upset plot}
#| lightbox: true

log_info('DE Summary computation...')
listInput <- list()
df_summary = as.data.frame(matrix(ncol=3, nrow=0))
i <- 1 
for (cmp in names(res_de)){
  cmp_label <- gsub(params$contrast,'',cmp )
  total =  dim(res_de[[cmp]]$toptable %>% filter(differential_expressed =='UP' |  differential_expressed =='DOWN') )[1]
  up  = dim(res_de[[cmp]]$toptable %>% filter(differential_expressed =='UP') )[1]
  down  =   dim(res_de[[cmp]]$toptable %>% filter(differential_expressed =='DOWN') )[1]
  log_info(cmp)
  log_info(total)
  #df_summary[[cmp_label]]  <- c(total, up,down)
  df_summary[i,1]  <- total
  df_summary[i,2]  <- up
  df_summary[i,3]  <- down

  if (total >  0){
    listInput[[cmp_label]] <- res_de[[cmp]]$toptable %>% filter(differential_expressed =='UP' |  differential_expressed =='DOWN') %>% pull(Protein.Names)
  }
  i <- i+1
}
rownames(df_summary) <- names(res_de) 
colnames(df_summary) <- c('Total DE','Up DE','Down DE')
    ## visual table summary
    DT::datatable( df_summary  ,
             extensions = c('FixedColumns', 'Scroller'),
             options = list(fixedColumns = TRUE, scrollY = 400, scrollX = TRUE,
                            scroller = TRUE, dom = 'Bfrtip', autoWidth = TRUE
             ),
     ) %>% formatStyle(
   columns= c('Total DE','Up DE','Down DE'),
   textAlign = 'center')
    # if there's 1 or 2 elements inlistInput and one of them is empty, skip uspset plot
 if (length(listInput)>=2 ) {
    
    print (upset(fromList(listInput), order.by = "freq", nsets = length(listInput) ) )
     
      pdf(file= paste0(file.path( params$folder_prj,"Result","Contrast"),'Upset_plot.pdf'),paper="a4")
       
     print( upset(fromList(listInput),  nsets = length(listInput) , order.by = "freq",text.scale= 1,
                  empty.intersections = "on")) 
     
      invisible(dev.off())
      
      
 
}
    

```

## Session Info

```{r session_info }
sessionInfo()
```
